---
title: 'A FODO Cell'
teaching: 10
exercises: 20
---

:::::::::::::::::::::::::::::::::::::: questions 

How to simulate a particle beam travelling through a FODO cell 🧲?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

Learn how to setup, run, and visualize a particle 
beam simulation through a FODO (Focusing-Defocusing) cell with WarpX

::::::::::::::::::::::::::::::::::::::::::::::::

## Setup

In this example we will simulate a particle beam travelling through a FODO cell.
A **FODO cell** is a periodic focusing structure used in particle accelerators, consisting of alternating **F**ocusing (F) and **D**efocusing (D) quadrupole magnets, with **O** (drift) sections in between.
The FODO cell is one of the most fundamental building blocks in accelerator physics, providing transverse focusing to keep particle beams confined as they travel through the accelerator.
The alternating focusing and defocusing quadrupoles create a net focusing effect in both transverse planes, allowing the beam to be transported over long distances while maintaining its size.

### WarpX for Beam Dynamics

WarpX is a Particle-in-Cell (PIC) code that can track particles embedded in external fields. 
Two important features are:

* **Space charge effects**: 
You can turn space charge effects on or off using the appropriate input parameters. 
When enabled, WarpX solves the electromagnetic/static fields self-consistently from the particle distributions, 
allowing you to study how the beam's own fields affect its dynamics.

* **Arbitrary external fields**: 
WarpX allows you to define arbitrary external electromagnetic fields, 
either analytically or by loading them from files. 
This makes it possible to model complex accelerator elements like quadrupoles, dipoles, and other magnets.

::: caution

**WarpX is not necessarily the best tool for all beam dynamics simulations**. 
If space charge effects are negligible and you're primarily interested in the linear beam optics, 
there are more specialized and computationally efficient tools available, such as ImpactX.

:::

### WarpX vs ImpactX

**ImpactX** is another code from the same ecosystem, the Beam, Plasma & Accelerator Simulation Toolkit ([**BLAST**][blast]), that is specifically designed for beam dynamics simulations. 
You can find the [ImpactX documentation][impactx] and [repository][impactx-github] for more information.
Here are the key differences:

* **ImpactX** is optimized for beam dynamics applications and uses envelope tracking or symplectic particle tracking methods. 
It is typically much faster than WarpX for cases where space charge effects are weak or can be approximated.

A fundamental distinction between these codes is their evolution variable:

* **WarpX** is a **t-based** (time-based) code: 
It evolves the system in time by solving Maxwell's or Poisson's equations on a spatial grid and advancing particles and fields self-consistently in time steps. 
This makes it suitable for:
  - Strong space charge effects that require self-consistent field solutions
  - Wakefield effects that need to be captured
  - Full electromagnetic field solutions
  - Plasma-beam interactions or other complex collective effects
  - Cases where the full time evolution of fields is important

* **ImpactX** is a **z-based** (position-based) code: 
It tracks particles along the beamline (z-direction) through accelerator elements, advancing the beam's state as it propagates through the lattice. 
This approach is more efficient for:
  - Space charge effects that are negligible or can be approximated
  - Fast simulations for optics design and optimization
  - Linear beam dynamics studies
  - Envelope models for rapid prototyping
  - Cases where you're primarily interested in the beam's evolution through the accelerator structure rather than detailed time-dependent field evolution

For this FODO cell tutorial, we'll use WarpX to demonstrate how to set up a beam dynamics simulation with the option to include or exclude space charge effects.

::: callout
## Equivalent ImpactX Example

If you're interested in seeing how a similar FODO cell simulation can be set up using ImpactX, 
check out the [ImpactX FODO cell example with 2D space charge using envelope tracking][impactx-fodo-example]. 
This example demonstrates the envelope tracking approach.
:::


Whenever you need to prepare an input file, [this is where you want to go](https://warpx.readthedocs.io/en/latest/usage/parameters.html).

:::::::::::::::::::::::::::::::::::::::::: spoiler

### Let's take a look at the input file

```{r engine='bash', echo=FALSE, comment=''}
cat ./files/inputs_3d_fodo_cell.txt
```
::::::::::::::::::::::::::::::::::::::::::::::::::


Some notable details:

*  [Details about the simulation setup will be added here]

*  [Information about particle initialization will be added here]

*  [Details about the magnetic field configuration will be added here]

*  [Information about diagnostics will be added here]

## Run 

First things first. 
Create a new folder where you copy the [input file](./files/inputs_3d_fodo_cell.txt). 
The simulation should be small enough that you can run it in serial with the Conda installation of WarpX.

::: spoiler

## Let's run in serial

```bash
warpx.3d inputs_3d_fodo_cell.txt
```

Just like that! 💃
Note that with Conda's WarpX you can run this anywhere in your filesystem (provided that you copied there the input of course)
because WarpX's executables are in your `$PATH`.
:::



If you want to make the simulation faster and/or bigger, then you should run with the parallel version of WarpX.
The optimal setup to run the simulation depends on your hardware. 
This is an example that should work on many common laptops, even though it might not be ideal.


::: spoiler

## Let's run in parallel

This is just one way of doing it!

```bash
export OMP_NUM_THREADS=2
mpirun -np 4 <path/to/your/build/bin/warpx.3d> inputs_3d_fodo_cell.txt 
```
:::

::: testimonial

With the default parameters, it may take a while on a laptop.
If you have access to a GPU, you can experience the speed-up yourself.
My experiments: 35 seconds on an NVIDIA A100 GPU, more than 1 hour on 4 cores of my 12th Gen Intel(R) Core(TM) i9-12900H.

:::


## Visualize

### With Python 🐍
Now that we have the results, we can analyze them using Python.  
We will use the [openPMD-viewer][openpmd-viewer] library to grab the data that the simulation produced in `openPMD` format. 
Here you can find [a few great tutorials on how to use the viewer](https://openpmd-viewer.readthedocs.io/en/latest/tutorials/tutorials.html).
If you feel nerdy and/or you need to deal with the data in parallel workflows, you can use the [openPMD-api][openpmd-api].

As an example for the FODO cell simulation, we have developed a simple Jupyter notebook where we retrieve 
the beam particle positions and analyze the beam dynamics through the cell.

:::::::::::::::::::::::::::::::::::::::::: spoiler

## Let's take a look at the Jupyter notebook

```{r, results='asis', echo=FALSE}
cat('<iframe src="https://nbviewer.org/github/BLAST-WarpX/warpx-tutorials/blob/main/episodes/files/analysis_3d_fodo_cell.ipynb" 
           width="100%" height="800" style="border:none;">
     </iframe>')
```
::::::::::::::::::::::::::::::::::::::::::::::::::

You can [download the notebook](./files/analysis_3d_fodo_cell.ipynb) and try it yourself.
Remember to either run the notebook from the simulation directory or change the corresponding path in the notebook.


::::::::::::::::::::::::::::::::::::: keypoints 
 🎯 Particle tracking follows individual particles, while envelope tracking follows the beam envelope - particle tracking is more detailed but computationally expensive.

 🔬 A FODO cell is a periodic focusing structure with alternating focusing and defocusing quadrupole magnets.

 ⚡ WarpX allows you to turn space charge effects on/off and define arbitrary external fields.

 🚀 WarpX is a full PIC code best suited for strong space charge effects and complex electromagnetic interactions. 
 For cases with negligible space charge, **ImpactX** may be a more efficient choice.

 🔍 The [**documentation**][warpx-readthedocs] is the first place to look for answers, 
 otherwise check out our [**issues**][warpx-issues] and [**discussions**][warpx-discussions] and ask there.  

 📷 To analyze and visualize the simulation results in [**openPMD**][openpmd] format, 
 you can use the [**openPMD-viewer**][openpmd-viewer] library for Python.

::::::::::::::::::::::::::::::::::::::::::::::::
